---
title: "BombusCanadaSpatialAnalysis-Data Preprocess"
output:
  html_document:
    df_print: paged
---

Data preprocess for Bombus of Canada spatial analysis.

```{r}
# Load libraries.
library(readr)
library(dplyr)
library(ggplot2)
library(spatstat)
library(sp)
library(sf)
library(spatstat.geom)
library(RColorBrewer)
```

## Read dataset
### 1. Read Bombus occurrence data. 
```{r}
# Read occurrence data
bombus_data <- read_tsv("../data/0008996-250402121839773/occurrence.txt")

# Check data.
dim(bombus_data)
```

Filter to entries with coordinates and genus `Bombus`:

```{r}
bombus_filtered <- bombus_data %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude), genus == "Bombus")
```

Removed `r dim(bombus_data)[1] - dim(bombus_filtered)[1]` records without coordinates and not belong to `Bombus`.

Totally, there are `r dim(bombus_filtered)[1]` bombus records. 

Check `specificEpithet`:

> Definition -- Specific Epithet
>
> In biological taxonomy, a specific epithet is the second part of the scientific name (binomial name) of a species. It identifies the species within a given genus.


```{r}
unique_specific_epithet_all <- unique(bombus_filtered$specificEpithet)
```

There are `r length(unique_specific_epithet_all)` specific species in North America.

### 2. Read BC Covariates.
```{r}
load('../data/BC_Covariates.Rda')

data_covariates <- DATA
```

### 3. Read BC Park
```{r}
load('../data/BC_Parks.Rda')
data_parks <- DATA

# Create the parks ppp object
ppp_parks <- ppp(
  x = data_parks$Parks$X, 
  y = data_parks$Parks$Y, 
  window = as.owin(data_parks$Window))

# Add region information as marks
marks(ppp_parks) <- data_parks$Parks$Region
```

## Explore the Covariate Data

```{r}
sapply(data_covariates, class)
summary(data_covariates)
```

### Plot the Window and Image Class Objects
```{r fig.height=8, fig.width=6}

par(mfrow = c(3,2))
plot(data_covariates$Window)
plot(data_covariates$Elevation)
plot(data_covariates$Forest)
plot(data_covariates$HFI)
plot(data_covariates$Dist_Water)
```

## Filter Bumble Bee in BC Province
```{r}
bombus_bc <- subset(bombus_filtered, stateProvince == "British Columbia")

# View(bombus_bc)
# 3185 223
dim(bombus_bc)

bumble_bc_percent <- (dim(bombus_bc)[1] / dim(bombus_filtered)[1]) * 100
print(paste(bumble_bc_percent, "% of bumble bee sightings across Canada occurred in BC."))
```

There are **`r dim(bombus_bc)[1]`** bumble bee occurrences in BC provinces. We removed `r dim(bombus_filtered)[1] - dim(bombus_bc)[1]` records. 

### Check `specificEpithet`:
```{r}
unique_specific_epithet_bc <- unique(bombus_bc$specificEpithet)
```

There are `r length(unique_specific_epithet_bc)` specific species the province of British Columbia, accounting for `r length(unique_specific_epithet_bc)/length(unique_specific_epithet_all) * 100` % of the total found in North America.


### Plot BC province window
```{r}
# Convert the window to an owin object
sf_obj <- st_as_sf(data_covariates$Window)
window <- as.owin(sf_obj)
plot(window, main="BC Province Window")
```


### Plot bombus bc dataset
```{r}
# Check dataset columns.
# colnames(bombus_bc)
```


```{r}
# Convert the cleaned data frame to an sf object
data_sf <- st_as_sf(bombus_bc, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

# Define the target projection
projected_args <- "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000 +y_0=0 +datum=NAD83 +units=m +no_defs"

# Transform the coordinates
bombus_bc_projected <- st_transform(data_sf, crs = projected_args)
# View(bombus_bc_projected)
dim(bombus_bc_projected)

# Get coordinations
bombus_bc_projected_coords <- st_coordinates(bombus_bc_projected)
head(bombus_bc_projected_coords)

# Convert to ppp object.
ppp_bombus <- ppp(
  x = bombus_bc_projected_coords[, "X"],
  y = bombus_bc_projected_coords[, "Y"],
  window = window)

# Plot bombus data
plot(ppp_bombus,  main = 'Bumble Bee in BC Province')
```


### Remove points outside BC province and remove duplicates.
From the above plot warning, we can see there are 62 points outside the BC province window. There are also duplicates in our dataset.

```{r}
# Filter out points outside the window
inside_window <- bombus_bc_projected_coords[, "X"] >= window$xrange[1] & 
                 bombus_bc_projected_coords[, "X"] <= window$xrange[2] &
                 bombus_bc_projected_coords[, "Y"] >= window$yrange[1] &
                 bombus_bc_projected_coords[, "Y"] <= window$yrange[2]

marks_df <- data.frame(
  specificEpithet = bombus_bc_projected$specificEpithet[inside_window],
  locality        = bombus_bc_projected$locality[inside_window]
)

# Convert to ppp object
ppp_bombus <- ppp(x = bombus_bc_projected_coords[inside_window, "X"],
                 y = bombus_bc_projected_coords[inside_window, "Y"],
                 marks  = marks_df,
                 window = window)
```

Remove duplicates.
```{r}
# Remove duplicates
ppp_bombus <- ppp_bombus[!duplicated(ppp_bombus), ]
```

There are `r (dim(bombus_bc_projected)[1] - ppp_bombus$n)` duplicated points. Now, there are `r ppp_bombus$n` unique bomble bee datapoints left in BC province.

### Plot Specific Epithet vs. Bumble Bee Occurrences
```{r}
# Extract levels and number of classes
epithet_levels <- unique(marks(ppp_bombus)$specificEpithet)
n_classes <- length(epithet_levels)

# Define color palette
colors <- colorRampPalette(brewer.pal(12, "Set3"))(n_classes)

# Define shape values (repeats if n_classes > 26)
all_shapes <- 0:25
shape_values <- rep(all_shapes, length.out = n_classes)

# Plot
plot(ppp_bombus,
     which.marks = "specificEpithet",
     main        = "Point Pattern of Specific Epithet and Bumble Bee Occurrences",
     cols         = colors,
     pch         = shape_values,
     cex         = 0.8,
     legend      = FALSE
)

# Add a custom legend
legend(
  "topright", 
  legend = epithet_levels, 
  col    = colors, 
  pch    = shape_values, 
  cex    = 0.6, 
  ncol   = 2
)
```

### Add marks
Add `Distance`, `Dist Water`, `Elevation`, `Forest`, and `HFI` for marks.

```{r}
length_max <- length(dist)

# Distance -- 984 data points, no missing records.
dist <- nndist(ppp_bombus)

# Dist Water -- 984 data points, no missing records.
dist_water <- data_covariates$Dist_Water[ppp_bombus]

# Elevation -- 982 data points, 2 missing records, replace with NA.
elev <- data_covariates$Elevation[ppp_bombus]
length(elev) <- length_max

# Forest -- 963 data points, 19 missing points, replace with NA.
forest <- data_covariates$Forest[ppp_bombus]
length(forest) <- length_max

# HFI -- 950 data points, 34 missing points, replace with NA.
hfi_bombus <- data_covariates$HFI[ppp_bombus]
length(hfi_bombus) <- length_max

# Add marks.
marks(ppp_bombus)$Dist <- dist
marks(ppp_bombus)$DistWater <- dist_water
marks(ppp_bombus)$Elevation <- elev
marks(ppp_bombus)$Forest <- forest
marks(ppp_bombus)$HFI <- hfi_bombus

head(marks(ppp_bombus))
```
Plot
```{r}
# Plot distance mark.
plot(ppp_bombus, which.marks = "Dist", main = "Bomble Bee Distance")

# Plot dist water
plot(ppp_bombus, which.marks = "DistWater", main = "Dist Water")

# Plot elevation
ppp_bombus_elev_clean <- subset(ppp_bombus, !is.na(marks(ppp_bombus)$Elevation))
marks(ppp_bombus_elev_clean)$ElevationCut <- cut(marks(ppp_bombus_elev_clean)$Elevation, breaks = 4)
plot(ppp_bombus_elev_clean, which.marks = "ElevationCut", main = "Elevation")

# Plot Forest
ppp_bombus_forest_clean <- subset(ppp_bombus, !is.na(marks(ppp_bombus)$Forest))
plot(ppp_bombus_forest_clean, which.marks = "Forest", main = "Forest")

# Plot HFI
ppp_bombus_hfi_clean <- subset(ppp_bombus, !is.na(marks(ppp_bombus)$HFI))
plot(ppp_bombus_hfi_clean, which.marks = "HFI", main = "HFI")
```

## Park
Plot Park vs. Bumble bee occurrences.
```{r}
# Plot the point pattern & assign to variable for specific information
ppp_parks_plot <- plot(ppp_parks,
                       main = "Point Pattern of Parks and Bumble Bee Occurrences", 
                       col = "grey90", 
                       cols = brewer.pal(n = 5, name = "Set2"), 
                       pch = c(15, 19, 18, 17, 20), 
                       cex = 0.8, 
                       legend = FALSE
)

# Add a custom legend with a title 'Region'
legend("topright", 
       legend = c("North", "Ok", "South", "Tc", "West", "Bumble Bee"), 
       title = "Marks", 
       col = c(brewer.pal(n = 5, name = "Set2"), "yellow"), 
       pch = c(15, 19, 18, 17, 16, 20), 
       cex = 0.8
)

# Plot the bumble bee data points
plot(ppp_bombus, add = TRUE, cols = "black", pch = 20, cex = 0.6)
plot(ppp_bombus, add = TRUE, cols = "yellow", pch = 20, cex = 0.3)
```

### Save `ppp_bombus` object

```{r}
save(ppp_bombus, file = "../data/ppp_bombus.Rda")
```


