---
title: "BombusCanadaSpatialAnalysis-Poisson Point Process Models"
output:
  html_document:
    df_print: paged
---

Poisson Point Process Models for Bombus of Canada spatial analysis.

```{r}
# Load libraries.
library(readr)
library(dplyr)
library(ggplot2)
library(spatstat)
library(sp)
library(sf)
library(spatstat.geom)
library(RColorBrewer)
```

## Read dataset
```{r}
# Read ppp data
load('../data/ppp_bombus.Rda')

load('../data/BC_Covariates.Rda')
data_covariates <- DATA

load('../data/BC_Parks.Rda')
data_parks <- DATA
```

### Check for collinearity between covariates
```{r}
# Check for collinearity
cor_matrix_covariates <- cor.im(Elevation = data_covariates$Elevation, Forest = data_covariates$Forest, HFI = data_covariates$HFI, Dist_water = data_covariates$Dist_Water, use = "pairwise.complete.obs") # 
cor_matrix_covariates
```

```{r}
library(corrplot)

# Plot the correlation matrix using corrplot with values overlayed
cor_plot <- corrplot(cor_matrix_covariates, method = "number", type = "upper", order = "hclust", tl.col = "black", tl.srt = 45) 
```

### Fit a null model (intercept only)
```{r}
# Fit a null model
ppp_unmarked <- unmark(ppp_bombus)
fit_null <- ppm(ppp_unmarked ~ 1)
fit_null
```

### Fit a linear model
```{r}
# Iterate over each element in data_covariates
for (i in seq_along(data_covariates)) {
  # Check if the element is of class 'im'
  if (inherits(data_covariates[[i]], "im")) {
    # Replace NA and NaN values with zeros
    data_covariates[[i]]$v[is.na(data_covariates[[i]]$v) | is.nan(data_covariates[[i]]$v)] <- 0
  }
}

# Fit a linear model
fit_linear <- ppm(ppp_unmarked ~ Elevation + Forest + HFI + Dist_Water, data = data_covariates)
fit_linear
```

### # Fit the PPP model with squared terms
```{r}
# Fit the PPP model with squared terms
fit <- ppm(ppp_unmarked ~ Elevation + I(Elevation^2) + Forest + HFI + I(HFI^2) + Dist_Water, data = data_covariates)
fit
```

Adding squared terms of Forest and Dist_Water will lead to error.

### Model Comparison
```{r}
# AIC values
cat("Null model AIC:", AIC(fit_null), "\n")
cat("Linear model AIC:", AIC(fit_linear), "\n")
cat("Full model AIC:", AIC(fit), "\n")
```

Full Model has the lowest AIC value of 40387.41.

### Likelihood ratio test
```{r}
# Likelihood ratio test comparing null model and linear model
anova(fit_null, fit_linear, test = "LRT")
```

The linear model is better than the null model.

```{r}
# Likelihood ratio test comparing linear model and full model
anova(fit_linear, fit, test = "LRT")
```

The full model is better than the linear model.

### Visualize the fitted model
```{r}
# visualize the fitted model
plot(fit,
     se = FALSE,
     superimpose = FALSE,
     main = "Fitted Model")

plot(ppp_unmarked,
     pch = 16,
     cex = 0.5,
     cols = "white",
     add = TRUE)
plot(ppp_unmarked,
     pch = 16,
     cex = 0.3,
     cols = "black",
     add = TRUE)
```

### Plot the effects of the individual coefficients
```{r}
# Median forest
median_forest <- median(data_covariates$Forest)

# Median elevation
median_elev <- median(data_covariates$Elevation)

# Median HFI
median_HFI <- median(data_covariates$HFI)

# Median Dist_Water
median_dist_water <- median(data_covariates$Dist_Water)

## Forest effect
forest_effect <- effectfun(fit, "Forest", Elevation = median_elev, HFI = median_HFI, Dist_Water = median_dist_water, se.fit = T)

## Elevation effect
elev_effect <- effectfun(fit, "Elevation", Forest = median_forest, HFI = median_HFI, Dist_Water = median_dist_water, se.fit = T)

## HFI effect
HFI_effect <- effectfun(fit, "HFI", Dist_Water = median_dist_water, Elevation = median_elev, Forest = median_forest, se.fit = T)

# Dist_Water effect
dist_water_effect <- effectfun(fit, "Dist_Water", HFI = median_HFI, Elevation = median_elev, Forest = median_forest, se.fit = T)

# Plotting
par(mfrow = c(2,2))

# Plot the elevation effect 
plot(elev_effect,
     legend = FALSE,
     main = "Effect of Elevation")

# Plot the forest effect 
plot(forest_effect,
     legend = FALSE,
     main = "Effect of Forest")

# Plot the Dist_Water effect
plot(dist_water_effect,
     legend = FALSE,
     main = "Effect of Dist_Water")

# Plot the HFI effect
plot(HFI_effect,
     legend = FALSE,
     main = "Effect of HFI")
```

### Quadrat test
```{r}
# Quadrat test
quadrat.test(fit, nx=4, ny=2)
```
The small p value tells us that there’s a significant deviation from our model’s predictions.

### Calculate and plot the model residuals
```{r}
# calculate the residuals
res <- residuals(fit)

# visualize
plot(res, cols = "transparent", main = "Residual Plot of Full Model")
```

### Calculate the partial residuals
```{r}
# Calculate the partial residuals as a function of elevation
par_res_elev <- parres(fit, "Elevation")

# Calculate the relative intensity as a function of forest
par_res_forest <- parres(fit, "Forest")

# Calculate the partial residuals as a function of HFI
par_res_HFI <- parres(fit, "HFI")

# Calculate the relative intensity as a function of Dist_Water
par_res_dist_water <- parres(fit, "Dist_Water")

# Plotting
par(mfrow = c(2,2))
plot(par_res_elev,
     main = "",
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Level of Elevation (m)")
plot(par_res_forest,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Percentage of Forest Coverage")
plot(par_res_HFI,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Human Footprint Index (HFI)")
plot(par_res_dist_water,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Distance from Water (m)")
```

### Try to improve model
```{r}
# add interaction terms
fit2 <- ppm(ppp_unmarked ~ Elevation * Forest + Elevation * HFI + HFI * Dist_Water + I(Elevation^2) + I(HFI^2), data = data_covariates)
fit2
```

### Model Comparison
```{r}
# AIC
AIC(fit)
AIC(fit2)
```

The interaction model is better than the original one.

```{r}
# Likelihood ratio test
anova(fit, fit2, test = "LRT")
```

The interaction model is better than the original one.

### Visualize and compare the models
```{r}
# visualize the fitted model
par(mfrow = c(1,2))
plot(fit,
     se = FALSE,
     superimpose = FALSE,
     main = "Fitted Model without Interaction")

plot(ppp_unmarked,
     pch = 16,
     cex = 0.5,
     cols = "white",
     add = TRUE)
plot(ppp_unmarked,
     pch = 16,
     cex = 0.3,
     cols = "black",
     add = TRUE)

plot(fit2,
     se = FALSE,
     superimpose = FALSE,
     main = "Fitted Model with Interaction")

plot(ppp_unmarked,
     pch = 16,
     cex = 0.5,
     cols = "white",
     add = TRUE)
plot(ppp_unmarked,
     pch = 16,
     cex = 0.3,
     cols = "black",
     add = TRUE)
```

### Quadrat test
```{r}
# Quadrat test
quadrat.test(fit2, nx=4, ny=3)
```
The small p value tells us that there’s a significant deviation from our model’s predictions.

### Calculate and plot the model residuals
```{r}
# calculate the residuals
res2 <- residuals(fit2)

# visualize
par(mfrow = c(1,2))
plot(res, cols = "transparent", main = "Residual Plot of Full Model", zlim = c(-5e-8, 0))
plot(res2, cols = "transparent", main = "Residual Plot of Improved Model", zlim = c(-5e-8, 0))
```

The improved model has lower residuals in some areas.

